---
title: "Crime Analysis"
author: "Younes Sandi"
date: "11/17/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Start

Importing required libraries 

```{r}
options(warn = -1)
library(naniar)
library(anytime) #converting date-time
library(lubridate)
library(plyr) #importing multiple csv files and merging them
library(e1071)
library(dbscan)
library(MASS)
library(corrplot)
library(LiblineaR)
library(ggplot2)
library(ggmap)
library(dplyr)
library(gridExtra)
library(caret)
library(ggExtra)
library(readr)
library(janitor)


```


_____________________________________________________________________________
_____________________________________________________________________________

## Accessing Data, splitting train and test, random sampling for further analysis

_____________________________________________________________________________
_____________________________________________________________________________

```{r}
setwd("E:/Datasets/crime/crime dataset/Crime Analysis")
getwd()
list.files(pattern="*.csv")
```



```{r}
temp = list.files(pattern="*.csv")
dataset<- ldply(temp, read.csv, header=TRUE)
# dataset includes 97 subframe from 2010-12 to 2018-12
head(dataset)
dim(dataset)
# the dataset has 2841166 rows and 12 columns
```

```{r}
print(names(dataset))
```

```{r}
#preparing a dataset with appropriate features,
# removing "Crime.ID", "Reported.by", "Falls.within", "LSOA.code",
# "Last.outcome.category", and "Context"
drops<- c("Crime.ID", "Reported.by", "Falls.within", "LSOA.code",
          "Last.outcome.category","Context")
dataset<- dataset[, !(names(dataset) %in% drops)]
names(dataset)
```


```{r}
# the Dataset is too bulky, hence for the purpose of this analysis, 40% of data
# has been sampled randomely

set.seed(42)
rows<- sample(nrow(dataset), as.integer((nrow(dataset))*0.1))
sample_df<- dataset[rows, ]

# shuffling the dataset, and spliting the dataset into test(60%) and test(40%)
set.seed(47)
rows<- sample(nrow(sample_df), as.integer((nrow(sample_df))*0.4))
train<- sample_df[-rows, ]
test<- sample_df[rows, ]
write.csv(sample_df, "E:/Datasets/crime/crime dataset/Crime Analysis/data/sample_df.csv", row.names = T)
write.csv(train, "E:/Datasets/crime/crime dataset/Crime Analysis/data/train.csv", row.names = T)
write.csv(test, "E:/Datasets/crime/crime dataset/Crime Analysis/data/test.csv", row.names = T)
write.csv(train, "E:/Datasets/crime/crime dataset/Crime Analysis/data/train.csv", row.names = T)
write.csv(dataset, "E:/Datasets/crime/crime dataset/Crime Analysis/data/dataset.csv", row.names = T)

# random sampling 10% of the train dataset to reduce computation load of visualization

df<- train[sample(nrow(train), as.integer((nrow(train))*0.1)),]
head(df)
dim(df)
rm(train, test, dataset, sample_df)
write.csv(df, "E:/Datasets/crime/crime dataset/Crime Analysis/data/df.csv", row.names = T)
rm(df)
df<- read.csv("E:/Datasets/crime/crime dataset/Crime Analysis/data/df.csv", 
              header = T, stringsAsFactors = F, na.strings = c("", "NA", "#N/A"))
```

_____________________________________________________________________________
_____________________________________________________________________________

## Defining functions for analysis

_____________________________________________________________________________
_____________________________________________________________________________


```{r}
# functions to make variables from date info.
feat_crt<- function(dataframe){
  
  col_names<- names(dataframe)
  dataframe <- data.frame(matrix(unlist(dataframe),nrow=nrow(dataframe)),stringsAsFactors=FALSE)
  names(dataframe) <- col_names
  
  names(dataframe)[names(dataframe)== "X"]<- "Row_Number"
  names(dataframe)[names(dataframe)== "Month"]<- "Date"
  dataframe$Years <- year(anytime(dataframe$Date))
  dataframe$Months <- month(anytime(dataframe$Date))
  dataframe<-mutate(dataframe, Date_New=anydate(Date)) 
  dataframe$District<- gsub("On or near ", "", as.character(dataframe$LSOA.name))
  dataframe$District<- lapply(strsplit(dataframe$District," ",fixed=TRUE), "[", 1)
  dataframe$Location<- gsub("On or near ", "", as.character(dataframe$Location))
  row_dataframe= as.data.frame(dataframe$Row_Number)
  col_drops<- c("Row_Number", "LSOA.name", "Date")
  dataframe<- dataframe[, !(names(dataframe) %in% col_drops)]
  dataframe<-mutate(dataframe, District = sapply(District, toString))
  dataframe$Longitude<- as.numeric(dataframe$Longitude)
  dataframe$Latitude<- as.numeric(dataframe$Latitude)
  
  return(dataframe)
}


# functions to draw map of crime.
map_crime <- function(dataframe, crime) {
  filtered <- filter(dataframe, Crime.type %in% crime)
  plot <- ggmap(map, extent='device') + 
    geom_point(data=filtered, 
               aes(x=Longitude, y=Latitude, color=Category), alpha=0.1)
  return(plot)
}

plot_crime_month <- function(dataframe, crime, month) {
  filtered <- filter(dataframe, Crime.type %in% crime )
  filtered <- summarise(group_by(filtered, Months ,Years),
                        count = n())
  print(filtered)
  plot <- ggplot(data = filtered, 
                 aes(x = Months,y = count,color = Years)) +
    geom_point()
  
  return(plot)
}

# functions to make contour maps
map_contours <- function(data, alp) {
  p1 = ggmap(map, extent='device') + 
    geom_point(data=data, aes(x=Longitude, y=Latitude), alpha= alp) + 
    stat_density2d(aes(x = Longitude, y = Latitude,
                       fill = ..level.., alpha = ..level..),
                   size = 0.1, data = data, n=100,
                   geom = "polygon") +
    theme(legend.position="none")
  return(p1)
}

plot_marginals <- function(data) {
  p2 = ggplot(data=data, aes(x=Longitude, y=Latitude), alpha=0.1)+
    geom_point()
  p2 = ggMarginal(p2 + theme_gray(), type = "histogram",
                  fill = "steelblue", col = "darkblue")
  return(p2)
}
```

_____________________________________________________________________________
_____________________________________________________________________________

## Inspecting Dataframe and its variables

_____________________________________________________________________________
_____________________________________________________________________________
```{r}
dim(df)
names(df)
head(df)
str(df)
table(as.character(sapply(df, class)))
summary(df)
vis_miss(df)
sum(is.na(df)) #there is one missing value in the dataframe

#inspecting number of unique values
```

_____________________________________________________________________________
_____________________________________________________________________________

## Creating new features, changing data types, dropping junk columns, and renaming number of variables variables 

_____________________________________________________________________________
_____________________________________________________________________________


```{r}

df<- feat_crt(df)
table(as.character(sapply(df, class)))
head(df)

```


_____________________________________________________________________________
_____________________________________________________________________________

## Downloading required maps for the analysis

_____________________________________________________________________________
_____________________________________________________________________________

```{r}

ggmap::register_google()
#set.seed(20)
#map<-get_map("greater manchester, uk", zoom = 10)
#ggmap(map)

## defining coordination of Geater Manchester for the map

paste0("Longitude: max ", max(df$Longitude), " & min ", min(df$Longitude)) 
paste0("Longitude: max ", max(df$Latitude), " & min ", min(df$Latitude)) 
loc<- c(left=-2.780, right=-1.960, top=53.67, bottom=53.33)
map<-get_map(loc, color="bw")

```


_____________________________________________________________________________
_____________________________________________________________________________

## Univariate Analysis

_____________________________________________________________________________
_____________________________________________________________________________

```{r}
# Variable:(Longitude, Latitude)
# Data type: Numeric, Continues
# Overal Distribution of crime across the Greater Manchester

dist_map <- ggmap(map, extent='device') + 
  geom_point(data=df, aes(x=Longitude, y=Latitude),
             alpha = 1/20,color = "red")+
  scale_colour_brewer(type="qual")
dist_map

# Observation: the distribution of crime for a sample of ~41000 shows that the crime occurrence 
# is distributed across the province, however, number of crimes in Manchester, and
# populated towns is higher than the rural areas.


mg_plot <- plot_marginals(df)
mg_plot


ct_plot <- map_contours(df,.05)
ct_plot
# Observation: detailed contour maps show that these crimes are concentrated in Manchester city 
# more than other places


```

#_____________________________________________________________________________
```{r}
df[df$District=='NA',]
```


```{r}
# Variable: (District)
# Data type: Character

#df<- df[complete.cases(df),] 
df<- df[!(df$District=='NA'),]#removing missing values


cnt_gp_dst = df %>%
  group_by(District) %>%
  dplyr::summarize(n = dplyr::n())

cnt_gp_dst$District<-as.character(cnt_gp_dst$District)

cnt_gp_dst<- as.data.frame(cnt_gp_dst) %>% 
  transform(District= reorder(District,n))

ggplot(cnt_gp_dst) +
  geom_bar(aes(x=District, y=n, 
               color = District, fill = District),
           stat="identity")+
  theme(legend.position="None")+
  coord_flip()

per_gp_dst <- df %>% 
  group_by(District) %>% 
  summarise(count = n()) %>% 
  mutate(perc = count/sum(count)) 

per_gp_dst$District <- as.character(per_gp_dst$District)

per_gp_dst<- as.data.frame(per_gp_dst) %>% 
  transform(District= reorder(District,perc))

ggplot(per_gp_dst) +
  geom_bar(aes(x=District, y=perc, 
               color = District, fill = District),
           stat="identity")+
  theme(legend.position="None")+
  coord_flip()

# Observation: City of Manchester Accounts for roughly 30% of crimes in the province
# followed by Bolton (11%) and Wilgan (9%)

```
#_____________________________________________________________________________
```{r}
# Variable: (Crime.type)
# Data type: Character


per_gp_crm <- df %>% 
  group_by(Crime.type) %>% 
  summarise(count = n()) %>% 
  mutate(perc = count/sum(count)) 

per_gp_crm$Crime.type <- as.character(per_gp_crm$Crime.type)

per_gp_crm<- as.data.frame(per_gp_crm) %>% 
  transform(Crime.type= reorder(Crime.type,perc))

ggplot(per_gp_crm) +
  geom_bar(aes(x=Crime.type, y=perc, 
               color = Crime.type, fill = Crime.type),
           stat="identity")+
  theme(legend.position="None")+
  coord_flip()

# Observation: Violence and sexual offence accounts for more than 27% percent of
# crimes followed by Anti-social behavior (17%), and Public order (12%), the least
# occurrence belongs to Possession of weapons


```

#_____________________________________________________________________________
```{r}
# Variable: (Location)
# Data type: Character

per_gp_loc <- df %>% 
  group_by(Location) %>% 
  summarise(count = n()) %>% 
  mutate(perc = count/sum(count)) 

per_gp_loc$Location <- as.character(per_gp_loc$Location)

per_gp_loc<- as.data.frame(per_gp_loc) %>% 
  transform(Location= reorder(Location,perc))

per_gp_loc[(per_gp_loc$count)>100,]


ggplot(per_gp_loc[(per_gp_loc$count)>100,]) +
  geom_bar(aes(x=Location, y=perc, 
               color = Location, fill = Location),
           stat="identity")+
  theme(legend.position="None")+
  coord_flip()

paste0("These 6 Locations account for roughly ",
       round((sum(per_gp_loc[(per_gp_loc$count)>100, "count"])
              /sum(per_gp_loc$count))*100, 2),"% of whole crimes" )

# Observation: Parking areas, followed by shopping ares, and supermarkets are
# places with most occurrence of crimes


```

#_____________________________________________________________________________
```{r}
# Variable: (Years)
# Data type:Numeric, Discrete (interval) 

cnt_gp_yr = df %>%
  group_by(Years) %>%
  dplyr::summarize(n = dplyr::n())

cnt_gp_yr$Years<-as.character(cnt_gp_yr$Years)

#cnt_gp_yr<- as.data.frame(cnt_gp_yr) %>% 
  #transform(Years= reorder(Years,n))

ggplot(cnt_gp_yr) +
  geom_bar(aes(x=Years, y=n, 
               color = Years, fill = Years),
           stat="identity")+
  theme(legend.position="None")
  
ggplot(cnt_gp_yr) +
  geom_line(aes(x=Years, y=n, group=1), color='red3')
# Observation: Since this dataset does not include whole year information for
# 2017 and 2019, further analysis of this feature is not possible

```

#_____________________________________________________________________________
```{r}
# Variable: (Months)
# Data type:Numeric, Discrete (interval) 

per_gp_mth <- df %>% 
  group_by(Months) %>% 
  summarise(count = n()) %>% 
  mutate(perc = count/sum(count)) 

per_gp_mth$Months <- as.character(per_gp_mth$Months)

#per_gp_mth<- as.data.frame(per_gp_mth) %>% 
  #transform(Months= reorder(Months,perc))
per_gp_mth$Months <- factor(per_gp_mth$Months, levels=c(seq(12)))
ggplot(per_gp_mth) +
  geom_bar(aes(x=Months, y=perc, 
               color = Months, fill = Months),
           stat="identity")+
  theme(legend.position="None")
ggplot(per_gp_mth) +
  geom_line(aes(x=Months, y=count, group=1), color='red3')  

# Observation: Since this dataset does not include whole year information for
# 2017 and 2019, further analysis of this feature is not possible


```

#_____________________________________________________________________________

```{r}
# Variable: (Date)
# Data type:Numeric, Discrete (interval) 

per_gp_dt <- df %>% 
  group_by(Date_New) %>% 
  summarise(count = n()) %>% 
  mutate(perc = count/sum(count)) 

per_gp_dt$Date_New <- as.character(per_gp_dt$Date_New)

per_gp_dt<- as.data.frame(per_gp_dt) 
  

ggplot(per_gp_dt) +
  geom_line(aes(y=count, x=Date_New, group=1), 
               color = 'red3',stat="identity")+
  theme(legend.position="None", axis.text.x = element_text(angle = 90, size = 5))

# Observation: Number of crimes varies across the time and no seasonality can
# be inferred from this plot, further investigation in terms of trend and seasonality
# is possible by investigating bivariate and multivariate analysis in next section

norm_gp_dt <- df %>% 
  group_by(Date_New) %>% 
  summarise(count = n()) %>%
  mutate(norm_count = (count-mean(count))/sd(count))

ggplot(norm_gp_dt) +
  geom_line(aes(y=norm_count, x=Date_New, group=1), 
               color = 'red3',stat="identity")+
  theme(legend.position="None", axis.text.x = element_text(angle = 90, size = 5))
```
_____________________________________________________________________________
_____________________________________________________________________________

## Bivariate Analysis

_____________________________________________________________________________
_____________________________________________________________________________


